### 2022-03-26 ho-techniques/ho-techniques.tex
biblio

In 2019 we tested for the first time if the idea of gracefully extending first-order provers to
higher-order logic really improves the state of the art. We
implemented \lsup{} \cite{bbtvw-21-sup-lam} in Zipperposition~1.5 with basic
heuristics and rudimentary extensions of the calculus to deal with Booleans. It
finished third at that year's higher-order division of CASC competition
\cite{gs-19-casc27}, 12~percentage points behind the winner, the tableau prover
Satallax 3.4 \cite{cb-2013-satallax}.

@@

Cooperation with efficient off-the-shelf first-order theorem provers is an
essential feature of higher-order theorem provers such as Leo-III
\cite[Sect.~4.4]{as-18-phd} and Satallax \cite{cb-2013-satallax}.
% but also HOLyHammer \cite{ku-15-holyhammer} and Sledgehammer \cite{bn-10-sh}.
Those provers invoke first-order backends repeatedly
during a proof attempt and spend a substantial amount of time in backend
collaboration. Since \lsup{} generalizes a highly efficient
first-order calculus, we expect that future efficient \lsup{}
implementations will not benefit much from backends.
Nevertheless, experimental provers such
as Zipperposition can still gain a lot. We present some
techniques for controlling the use of backends.

@@

%The raw evaluation data for the previous experiments show that
\looseness=-1
Different choices of
parameters lead to noticeably different sets of proved problems. In an attempt
to use Zipperposition~2 to its full potential, we have created a portfolio mode
that runs up to 50 configurations in parallel during the allotted time. The portfolio
was designed to solve as many problems as possible from the TPTP benchmark set. To
provide some context, we compare Zipperposition~2 with the following versions of
all other higher-order provers that competed at CASC in 2020:\ CVC4 1.9
\cite{cbetal-11-cvc4}, Leo-III 1.5.6 \cite{sb-21-leo3}, Satallax 3.5
\cite{cb-2013-satallax}, and Vampire 4.5.1 \cite{br-20-full-sup-w-combs}.
The provers were run using the same parameters as in CASC, but with updated
executables.
Note that
Vampire's higher-order schedule is optimized for running on a single core.
We also include Ehoh~2.7, the first version of this prover to syntactically support
full higher-order logic, including $\lambda$-abstractions.
Semantically, Ehoh~2.7 is arguably the weakest among the listed provers:
It simply performs $o$-RW rewriting described in Sect.~\ref{sec:ho-tech:preprocessing} followed by
$\lambda$-lifting before it applies \relax{\lfsup{}} \cite{bbcw-21-lfho}
on the preprocessed problem. It is a conservative extension of Ehoh presented in Chapter~\ref{ch:ehoh},
and serves as a baseline for comparison with $\lambda$E introduced in Chapter~\ref{ch:ehoh2}.

>>>>

In 2019 we tested for the first time if the idea of gracefully extending first-order provers to
higher-order logic really improves the state of the art. We
implemented \lsup{} \cite{bbtvw-21-sup-lam} in Zipperposition~1.5 with basic
heuristics and rudimentary extensions of the calculus to deal with Booleans. It
finished third at that year's higher-order division of CASC competition
\cite{gs-19-casc27}, 12~percentage points behind the winner, the tableau prover
Satallax 3.4 \cite{cb-12-satallax}.

@@

Cooperation with efficient off-the-shelf first-order theorem provers is an
essential feature of higher-order theorem provers such as Leo-III
\cite[Sect.~4.4]{as-18-phd} and Satallax \cite{cb-12-satallax}.
% but also HOLyHammer \cite{ku-15-holyhammer} and Sledgehammer \cite{bn-10-sh}.
Those provers invoke first-order backends repeatedly
during a proof attempt and spend a substantial amount of time in backend
collaboration. Since \lsup{} generalizes a highly efficient
first-order calculus, we expect that future efficient \lsup{}
implementations will not benefit much from backends.
Nevertheless, experimental provers such
as Zipperposition can still gain a lot. We present some
techniques for controlling the use of backends.

@@

%The raw evaluation data for the previous experiments show that
\looseness=-1
Different choices of
parameters lead to noticeably different sets of proved problems. In an attempt
to use Zipperposition~2 to its full potential, we have created a portfolio mode
that runs up to 50 configurations in parallel during the allotted time. The portfolio
was designed to solve as many problems as possible from the TPTP benchmark set. To
provide some context, we compare Zipperposition~2 with the following versions of
all other higher-order provers that competed at CASC in 2020:\ CVC4 1.9
\cite{cbetal-11-cvc4}, Leo-III 1.5.6 \cite{sb-21-leo3}, Satallax 3.5
\cite{cb-12-satallax}, and Vampire 4.5.1 \cite{br-20-full-sup-w-combs}.
The provers were run using the same parameters as in CASC, but with updated
executables.
Note that
Vampire's higher-order schedule is optimized for running on a single core.
We also include Ehoh~2.7, the first version of this prover to syntactically support
full higher-order logic, including $\lambda$-abstractions.
Semantically, Ehoh~2.7 is arguably the weakest among the listed provers:
It simply performs $o$-RW rewriting described in Sect.~\ref{sec:ho-tech:preprocessing} followed by
$\lambda$-lifting before it applies \relax{\lfsup{}} \cite{bbcw-21-lfho}
on the preprocessed problem. It is a conservative extension of Ehoh presented in Chapter~\ref{ch:ehoh},
and serves as a baseline for comparison with $\lambda$E introduced in Chapter~\ref{ch:ehoh2}.

### 2022-03-26 introduction/introduction.tex
biblio

Despite this negative result, the prospect of automatically proving theorems of
first-order logic remained too enticing. In 1960, Davis and Putnam
\cite{dp-1960-dpll} described a procedure for checking validity of a
first-order formula. As first-order logic is undecidable, this procedure
terminated only on valid formulas. Even though it was efficient enough to prove
only simple formulas, it was an impressive achievement for the time.

@@

  Together with Alexander Bentkamp and other colleagues we developed
  a complete superposition calculus for the second phase described above
  \cite{bbtvw-21-sup-lam}. Due to the complexity of the calculus and many design
  decisions that had to be evaluated in a short amount of time we decided to
  implement the calculus in Zipperposition \cite{sc-15-simon-phd,sc-supind-17}. This prover is less efficient than
  E, but it is implemented in OCaml and it was designed to make prototyping
  ideas and experimenting with calculi much easier. In 2019 we entered the
  higher-order division of CASC theorem proving competition with Zipperposition
  \cite{gs-19-casc27}. It finished third, 1 percent point behind Leo-III
  \cite{sb-21-leo3}  and 12 percent points behind  Satallax
  \cite{cb-12-satallax}.

>>>>

Despite this negative result, the prospect of automatically proving theorems of
first-order logic remained too enticing. In 1960, Davis and Putnam
\cite{dp-60-dp} described a procedure for checking validity of a
first-order formula. As first-order logic is undecidable, this procedure
terminated only on valid formulas. Even though it was efficient enough to prove
only simple formulas, it was an impressive achievement for the time.

@@

  Together with Alexander Bentkamp and other colleagues we developed
  a complete superposition calculus for the second phase described above
  \cite{bbtvw-21-sup-lam}. Due to the complexity of the calculus and many design
  decisions that had to be evaluated in a short amount of time we decided to
  implement the calculus in Zipperposition \cite{sc-15-simon-phd,sc-supind-17}. This prover is less efficient than
  E, but it is implemented in OCaml and it was designed to make prototyping
  ideas and experimenting with calculi much easier. In 2019 we entered the
  higher-order division of CASC theorem proving competition~\cite{gs-19-casc27} with Zipperposition. 
  It finished third, one percent point behind Leo-III
  \cite{sb-21-leo3}  and 12 percent points behind  Satallax
  \cite{cb-12-satallax}.

### 2022-03-26 unif/unif.tex
more
biblio

\section{Implementation}
\label{sec:unif:implementation}
Zipperposition \cite{sc-15-simon-phd, sc-supind-17} is an
open-source\footnote{\url{https://github.com/sneeuwballen/zipperposition}}
theorem prover written in OCaml.
It is a versatile testbed for prototyping extensions to
superposition-based theorem provers.
It was initially
designed as a prover for polymorphic first-order logic
and then extended to higher-order logic.
A recent addition is a complete mode for
Boolean-free higher-order logic \cite{bbtvw-21-sup-lam},
which depends on a unification procedure that can enumerate a CSU. 
We implemented our procedure in Zipperposition.

>>>>

\section{Implementation}
\label{sec:unif:implementation}
As described in Sect.~\ref{sec:pre:theorem-provers}, Zipperposition \cite{sc-15-simon-phd, sc-supind-17} is an
open-source\footnote{\url{https://github.com/sneeuwballen/zipperposition}}
theorem prover written in OCaml.
It is a versatile testbed for prototyping extensions to
superposition-based theorem provers.
It was initially designed as a prover for rank-1 polymorphic \cite{bp-13-tff1} first-order logic 
and then extended to higher-order logic.
A recent addition is a complete mode for
Boolean-free higher-order logic \cite{bbtvw-21-sup-lam},
which depends on a unification procedure that can enumerate a CSU. 
We implemented our procedure in Zipperposition.
